\documentclass[11pt]{article}

\usepackage{keylogo}
\usepackage{a4wide}

\newcommand{\java}[1]{\mbox{\textsf{#1}}}

\begin{document}
\title{Inductive Proof of a simple while loop}
\author{Andreas Roth}
\date{April 4, 2003}
\maketitle

\section*{Proof obligation}
\[<\java{int i;}> \forall il:\java{int}. \{i:=il\}(\java{i}\geq 0 \rightarrow ( <\java{while (i$>$0) i\,-\,-;}> \java{i}=0))\]

\noindent Decrementing a non-negative variable as long as the counter is
positive sets the variable to zero.

\section*{Proof}

Induction over $\java{il}$.

\section*{Proof with \KeY\footnote{tested with prover version 0.528}}

\begin{enumerate}
	\item Load input file \texttt{decrToZero.key} into the standalone
	prover. Natural number LDT setting is \emph{Arithmetic semantics
	(ignoring overflow)}. Update simplification setting is \emph{Use
	simultaneous updates} and \emph{Delete simple updates} and \emph{Delete
	unused updates}.  

	\item Eliminate the java block that declares the variable \java{i}. Do
	this by the taclets eliminate\_variable\_declaration and
	empty\_diamond.  

	\item Select taclet \textsf{int\_induction} (available on sequent
	arrow) \item Instantiate schema variables 

	\begin{itemize} 

		\item \textsf{b} to \\ \texttt{(geq(il,0))-> \{i:=il\}<\{ while
( i>0 )  i--;\}>i = 0}			
(\emph{Drag and Drop} might help.)
		\item \textsf{nv} 
			to \texttt{il} (the induction variable) 
		\end{itemize}
		and click \textsf{OK} in the instantiation dialog.
	\item Run all heuristics.
	\item Induction base case (\emph{Case 1}):
		Apply \textsf{while\_right} on the first formula in the
		succedent and apply all heuristics, which close the base case.
	\item Induction step (\emph{Case 2}) \footnote{Never include
\emph{heuristics `loop\_expand'} when doing induction!}:
		
		Two new goals have to be closed:
		\begin{itemize}
			\item The first one: Apply \textsf{while\_right} on the
				first formula in the 
				succedent and apply all heuristics.

				This can be closed by proving the contradiction
				of the formulas in the antecedent. For example
				by running the decision procedure
				``simplify''. Select it from the menu at the
				sequent arrow.

			\item The second one: Again, apply
			\textsf{while\_right} on the first formula in the
			succedent and apply all heuristics.

			We can close by rewriting the first DL formula in the
			succedent to equal the DL formula in the antecedent. To
			do that, the doubled Java block has to be eliminated
			using the taclet \textsf{elim\_double\_block}.

			Apply all heuristics then.

		\end{itemize} \item Show original goal with induction
		hypothesis (\emph{Case 3}): Apply \textsf{all\_left} to the
		formula in the antecedent (the induction hypothesis and
		instantiate the schema variable $t$ by the variable on the
		right hand side of the (only) update in the
		succedent. (Currently this is an epsilon term which you can
		utilise by the abbreviation, e.g.~``@eps3''. Apply all
		heuristics \emph{except from the heuristics `loop\_expand'}.
\end{enumerate}
\end{document}